Full Dissertation

### 2024-04-11 Sections/full.qmd
finish intro

---
title: Full Dissertation
subtitle: "The Role of Variability in Learning Transfer: A Similarity-Based Computational Approach"
date: 04/04/24
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
cache: true
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    include: false
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: false
    output: false
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
editor: 
  markdown: 
    wrap: 72
prefer-html: true
---

# Variability and Generalization

The factors that influence the generalization of learning are of considerable interest to both researchers exploring the human learning system and practitioners aiming to enhance the effectiveness of educational and training interventions. The present effort will focus specifically on the role of variability in the learning input. Variability manipulations typically regulate either the number of distinct instances presented to learners during training, or the dispersion of these instances. Such manipulations have been empirically demonstrated to affect subsequent generalization performance. This essay will offer an in-depth review of the extant literature on the influence of variability, spanning multiple relevant domains.

@@

Studies investigating the "benefits of variability" hypothesis usually assign participants to either a constant or varied group for the training stage of the experiment. Then, subjects in both groups complete an identical testing stage which often consists items/conditions seen during training, and novel items/conditions. If the varied group performs better in the testing stage, this is taken for evidence of the benefits of variability hypothesis. Even within this relatively straightforward between-groups design, researchers must navigate several crucial methodological choices, highlighted below:

1)  Variables Subject to Variation. In multidimensional tasks, researchers have the option to vary numerous variables. The experimenters must decide the specific dimension(s) across which variation will occur. For instance, in a projectile throwing accuracy task -- researchers might vary the distance from the target, the size of the target, the weight of the projectile. They might also vary a contextual variable not directly relevant to the task, but which will still be encoded by the subject on a trial by trial basis, e.g. the background color.

2)  Magnitude of Variation, relative to the control condition. The simplest comparison would be to compare a constant group who trains with 1 example/condition, against a varied group that trains from 2 examples/conditions. However, it is not uncommon in the literature for the varied condition to train from 3 or 4 conditions. For example, @catalanoDistantTransferCoincident1984a train varied subjects from 4 different velocities in their coincident timing task, and [@goodeSuperiorityVariableRepeated2008] have varied subjects' practice with 3 different variants (i.e. different letter scrambles of the same word) of an anagram for a given word, while their constant participants view the same variant 3 different times. Alternatively, rather than a constant vs. varied comparison, subjects in all conditions might experience a variety of training items, but with one group experiencing a greater number of unique items [@nosofskyModelguidedSearchOptimal2019].

3)  Locations within Task-Space. For tasks in which the stimuli or conditions fall within a continuous metric space, the experimenter must decide whether the varied instances are relatively close together (e.g. throwing a ball from a distance of 4 feet and 5 feet), or far apart (throwing from 4 feet and 20 feet). Spreading the varied training items further apart may be beneficial in terms of providing a more representative sample of the task space to the learner, however large distances may also result in significant differences in difficulty between the training examples, which can be a common confound in variability studies.

4)  Proximity of Testing to Training Conditions. Intuitively, the fairest form of comparison is to include testing conditions that are of an equivalent distance from both the varied and constant groups. However researchers might also attempt to demonstrate the benefits of variation as being sufficiently powerful to outperform constant training, even in cases where the constant group trained from a closer proximity to the testing conditions, or whose training conditions are identical to the testing conditions [@goodeSuperiorityVariableRepeated2008; @kerrSpecificVariedPractice1978].

## Variability Literature Review

An early and influential work on the influence of variability on category learning is that of @posnerGenesisAbstractIdeas1968. In an ambitious attempt to address the question of how category information is represented, the authors trained participants to categorize artificial dot patterns, manipulating whether learners were exposed to examples clustered close to the category prototypes (e.g. a low variability condition), or spread further away from the prototype (the varied-training group). It should be noted that both groups in this study were trained with the same number of unique instances and the manipulated difference was how spread out the instances were. The authors claim based on prior experiments using the same stimuli, that the training stimuli for the varied group were at least as far away from the testing stimuli as the training stimuli of the less-varied group. The authors interpreted their findings as evidence for the extraction of an abstraction or schema that is extracted and stored, and then over time becomes more likely to be the reference point from which generalization occurs, given that specific instances are thought to decay at a faster rate than prototypes or schema. The Posner and Keele study has been extremely influential and continues to be cited in contemporary research as clear evidence that schema abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of "Schema Theory of Motor Learning" [@schmidtSchemaTheoryDiscrete1975], which in turn influenced decades of investigations on the potential benefits of varied training in motor skill learning. However, the classic Posner & Keele study despite being far more carefully designed than many subsequent studies, and despite being a relative rarity in explicitly discussing and attempting to control for potential confounds of similarity between groups, may nevertheless be emblematic of a common issue in many investigations of the effects of varied training on learning. The problem with Posner & Keele's conclusion was demonstrated clearly almost 3 decades later [@palmeriCentralTendenciesExtreme2001], when researchers conducting a near replication of the original study also collected similarity judgements following training and performed multidimensional scaling analysis. Rather than being in the middle of the training stimuli as was the case in the physical stimuli space, the psychological representation of the prototype was shown reside at an extreme point, and generalization patterns by participants that would have seemed to warrant the learning of a prototype were then easily accounted for with only the assumption that the participants encoded instances. One of the primary concerns of the present paper is that many of the studies which purport to explain the benefits of variation via prototypes, schemas, or other abstractions, are often overlooking the potential of instance based similarity accounts.

## Motor skill learning

Training variation has also been shown to promote transfer in motor learning. Much of this research has been influenced by the work of [@schmidtSchemaTheoryDiscrete1975], who proposed a schema-based account of motor learning as an attempt to address the longstanding problem of how novel movements are produced. Schema theory presumes a priori that learners possess general motor programs for classes of movements, such as an underhand throw. When called up for use, such programs must be parameterized, as well as schema rules that determine how a motor program is parameterized or scaled for a particular movement. Schema theory predicts that varied training results in the formation of a more general schema-rule, which can allow for transfer to novel movements within a given movement class, such as an underhand throw (though it is agnostic to the development of the movement classes themselves). Experiments that test this hypothesis are often designed to compare the transfer performance of a constant-trained group against that of a varied-trained group. Both groups train on the same task, but the varied group practices with multiple instances along some task-relevant dimension that remains invariant for the constant group. For example, investigators might train two groups of participants to throw a projectile at a target, with a constant group that throws from a single location, and a varied group that throws from multiple locations. Both groups are then tested from novel locations.

One of the earliest, and still often cited investigations of Schmidt's benefits of variability hypothesis was the work of @kerrSpecificVariedPractice1978. Two groups of children, aged 8 and 12, were assigned to either constant or varied training of a bean bag throwing task. The constant group practiced throwing a bean-bag at a small target placed 3 feet in front of them, and the varied group practiced throwing from a distance of both 2 feet and 4 feet. Participants were blindfolded and unable to see the target while making each throw but would receive feedback by looking at where the beanbag had landed in between each training trial. 12 weeks later, all of the children were given a final test from a distance of 3 feet which was novel for the varied participants and repeated for the constant participants. Participants were also blindfolded for testing and did not receive trial by trial feedback in this stage. However, at the halfway point of the testing stage they were allowed to see the landing location of the 4 beanbags they had thrown, and then completed the final 4 testing throws. In both age groups, participants performed significantly better in the varied condition than the constant condition, though the effect was larger for the younger, 8-year-old children. Although this design does not directly assess the hypothesis of varied training producing superior generalization to constant training (since the constant group is not tested from a novel position), it nevertheless offers a compelling example of the merits of varied practice.

On occasion the Kerr and Booth design may be nested within a larger experimental design. One such study that used a movement timing task, wherein subjects had to move their hand from a starting location, to a target location, attempting to arrive at the target location at specific time following the onset of a cue [@wrisbergVariabilityPracticeHypothesis1987]. This study utilized 4 different constant groups, and 3 varied groups, with one of the constant groups training under conditions identical to the testing conditions, and which were not trained on by any of the varied groups, e.g. the design of Kerr and Booth. However, in this case the varied group did not outperform the constant group. A more recent study attempting a slightly more direct replication of the original Kerr & Booth study [@willeyLongtermMotorLearning2018], having subjects throw beanbags at a target, with the varied group training from positions (5 and 9 feet) on either side of the constant group (7 feet). However, this study diverged from the original in that the participants were adults; they faced away from the target and threw the beanbag backwards over their bodies; they alternated using their right and left hands every 6 trials; and underwent a relatively extreme amount of training (20 sessions with 60 practice trials each, spread out over 5-7 weeks). Like @wrisbergVariabilityPracticeHypothesis1987, this study did not find a varied advantage from the constant training position, though the varied group did perform better at distances novel to both groups.

Some support for the Kerr and Booth findings was found with a relatively less common experimental task of training participants in hitting a projectile at a target with the use of a racket [@greenPracticeVariabilityTransfer1995a]. Varied participants trained with tennis, squash, badminton, and short-tennis rackets were compared against constant subjects trained with only a tennis racket. One of the testing conditions had subjects repeat the use of the tennis racket, which had been used on all 128 training trials for the constant group, and only 32 training trials for the varied group. Nevertheless, the varied group outperformed the constant group when using the tennis racket at testing, and also performed better in conditions with several novel racket lengths. Of course, this finding is less surprising than that of Kerr & Booth, given that varied subjects did have some prior exposure to the constant groups condition. This highlights an issue rarely discussed in the literature, of how much practice from an additional position might be necessary to induce benefits. Experimenters almost uniformly have varied participants train with an equivalent number of trials from each of their conditions

One of the few studies that has replicated the surprising result of varied outperforming constant, from the constant training condition, did so in the relatively distant domain of verbal manipulation [@goodeSuperiorityVariableRepeated2008]. All participants trained to solve anagrams of 40 different words ranging in length from 5 to 11 letters, with an anagram of each word repeated 3 times throughout training, for a total of 120 training trials. Although subjects in all conditions were exposed to the same 40 unique words (i.e. the solution to an anagram), participants in the varied group saw 3 different arrangements for each solution-word, such as DOLOF, FOLOD, and OOFLD for the solution word FLOOD, whereas constant subjects would train on three repetitions of LDOOF (spread evenly across training). Two different constant groups were used. Both constant groups trained with three repetitions of the same word scramble, but for constant group A, the testing phase consisted of the identical letter arrangement to that seen during training (e.g. LDOOF), whereas for constant group B, the testing phase consisted of a arrangement they had not seen during training, thus presenting them with a testing situation similar situation to the varied group. At the testing stage, the varied group outperformed both constant groups, a particularly impressive result, given that constant group A had 3 prior exposures to the word arrangement (i.e. the particular permutation of letters) which the varied group had not explicitly seen. However varied subjects in this study did not exhibit the typical decrement in the training phase typical of other varied manipulations in the literature, and actually achieved higher levels of anagram solving accuracy by the end of training than either of the constant groups -- solving 2 more anagrams on average than the constant group. This might suggest that for tasks of this nature where the learner can simply get stuck with a particular word scramble, repeated exposure to the identical scramble might be less helpful towards finding the solution than being given a different arrangement of the same letters. This contention is supported by the fact that constant group A, who was tested on the identical arrangement as they experienced during training, performed no better at testing than did constant group B, who had trained on a different arrangement of the same word solution -- further suggesting that there may not have been a strong identity advantage in this task.

Pitting varied against constant practice against each other on the home turf of the constant group provides a compelling argument for the benefits of varied training, as well as an interesting challenge for theoretical accounts that posit generalization to occur as some function of distance. However, despite its appeal this particular contrast is relatively uncommon in the literature. It is unclear whether this may be cause for concern over publication bias, or just researchers feeling the design is too risky. A far more common design is to have separate constant groups that each train exclusively from each of the conditions that the varied group encounters [@catalanoDistantTransferCoincident1984a; @chuaPracticeVariabilityPromotes2019; @newellVariabilityPracticeTransfer1976; @moxleySchemaVariabilityPractice1979; @mccrackenTestSchemaTheory1977], or for a single constant group to train from just one of the conditions experienced by the varied participants [@pigottMotorSchemaStructure1984; @rollerVariablePracticeLenses2001; @wrisbergTrainingProductionNovel1984; @wrisbergDevelopingCoincidentTiming1983]. A less common contrast places the constant group training in a region of the task space outside of the range of examples experienced by the varied group, but distinct from the transfer condition [@wrisbergVariabilityPracticeHypothesis1987; @wulfVariabilityPracticeImplicit1997].

Of particular relevant to the current essay is the early work of @catalanoDistantTransferCoincident1984a, as theirs was one of the earliest studies to investigate the influence of varied vs. constant training on multiple testing locations of graded distance from the training condition. Participants were trained on coincident timing task, in which subjects observe a series of lightbulbs turning on sequentially at a consistent rate and attempt to time a button response with the onset of the final bulb. The constant groups trained with a single velocity of either 5,7,9, or 11 mph, while the varied group trained from all 4 of these velocities. Participants were then assigned to one of four possible generalization conditions, all of which fell outside of the range of the varied training conditions -- 1, 3, 13 or 15 mph. As is often the case, the varied group performed worse during the training phase. In the testing phase, the general pattern was for all participants to perform worse as the testing conditions became further away from the training conditions, but since the drop off in performance as a function of distance was far less steep for the varied group, the authors suggested that varied training induced a decremented generalization gradient, such that the varied participants were less affected by the change between training and testing conditions.

## Category learning

In the category learning literature, the constant vs. varied comparison is much less suitable. Instead, researchers tend to compare a condition with many repetitions of a few items against condition with fewer repetitions of a wider array of exemplars. Much of the earlier work in this sub-area trained subjects on artificial categories, such as dot patterns [@homaCategoryBreadthAbstraction1976; @posnerGenesisAbstractIdeas1968], where more varied or distorted training examples were often shown to produce superior generalization when categorizing novel exemplars. More recently, researchers have also begun to utilize more realistic stimuli in their experiments. @wahlheimMetacognitiveJudgmentsRepetition2012 conducted one such study. In a within-participants design, participants were trained on bird categories with either high repetitions of a few exemplars, or few repetitions of many exemplars. Across four different experiments, which were conducted to address an unrelated question on metacognitive judgements, the researchers consistently found that participants generalized better to novel species following training with more unique exemplars (i.e. higher variability), while high repetition training produced significantly better performance categorizing the specific species they had trained on. A variability advantage was also found in the relatively complex domain of rock categorization [@nosofskyTestsExemplarmemoryModel2018]. For 10 different rock categories, participants were trained with either many repetitions of 3 unique examples of each category, or few repetitions of 9 unique examples, with an equal number of total training trials in each group (the design also included 2 other conditions less amenable to considering the impact of variation). The high-variability group, trained with 9 unique examples, showed significantly better generalization performance than the other conditions. Moreover, the pattern of results in this study could be nicely accounted for by an extended version of the Generalized Context Model.

The studies described thus far have studied the benefits of variability by exposing participants to a greater or lesser number of distinct examples during training. A distinct sub-literature within the category learning domain has focused much less on benefits derived from varied training, instead emphasizing how increased variability during the learning of a novel category influences how far the category boundary will then be generalized. The general approach is to train participants on examples from two categories, with the examples from one of the categories being more dispersed than the other. Participants are then tested with novel items located within ambiguous regions of the task space which allow the experimenters to assess whether the difference in variability influences how far participants generalize the category boundaries.

@cohenCategoryVariabilityExemplar2001 trained subjects on two categories, one with much more variability than the other. In experiment 1, a low variability category composed of 1 instance was compared against a high-variability category of 2 instances in one condition, and 7 instances in another. In experiment 2 both categories were composed of 3 instances, but for the low-variability group the instances were clustered close to each other, whereas the high-variability groups instances were spread much further apart. Participants were tested on an ambiguous novel instance that was located in between the two trained categories. Both experiments provided evidence that participants were much more likely to categorize the novel middle stimulus into a category with greater variation. Moreover, this effect was at odds with the predications of the baseline version of the GCM, thus providing some evidence that training variation may at least sometimes induce effects that cannot be entirely accounted for by exemplar-similarity accounts.

Further observations consonant with the results of @cohenCategoryVariabilityExemplar2001  have since been observed in numerous investigations [@hahnEffectsCategoryDiversity2005; @perlmanFurtherAttemptsClarify2012;
@sakamotoPuttingPsychologyBack2008; @hsuEffectsGenerativeDiscriminative2010]. The results of @sakamotoPuttingPsychologyBack2008 are noteworthy. They first reproduced the basic finding of participants being more likely to categorize an unknown middle stimulus into a training category with higher variability. In a second experiment, they held the variability between the two training categories constant and instead manipulated the training sequence, such that the examples of one category appeared in an ordered fashion, with very small changes from one example to the other (the stimuli were lines that varied only in length), whereas examples in the alternate category were shown in a random order and thus included larger jumps in the stimulus space from trial to trial. They found that the middle stimulus was more likely to be categorized into the category that had been learned with a random sequence, which was attributed to an increased perception of variability which resulted from the larger trial to trial discrepancies.

>>>>

---
title: Full Dissertation
subtitle: "The Role of Variability in Learning Transfer: A Similarity-Based Computational Approach"
date: 04/04/24
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
cache: true
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    #include: false
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: false
    output: false
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
editor: 
  markdown: 
    wrap: 72
prefer-html: true
---

The past century of research on human learning has produced ample evidence that although learners can improve at almost any task, such improvements are often specific to the trained task, with unreliable or even nonexistent transfer or generalization to novel tasks or conditions [@barnettWhenWhereWe2002; @dettermanCaseProsecutionTransfer1993]. Such generalization challenges are of noteworthy practical relevance, given that educators, trainers, and rehabilitators typically intend for their students to be able to apply what they have learned to new situations. It is therefore important to better understand the factors that influence generalization, and to develop cognitive models that can predict when generalization is likely to occur. Such characteristics have included training difficulty, spacing, temporal order, feedback schedules, and the primary focus of the current work - the variability of training examples.

@@

Varied training has been shown to influence learning in wide array of different tasks and domains, including categorization [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968; @nosofskyModelguidedSearchOptimal2019], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012], anagram completion [@goodeSuperiorityVariableRepeated2008], trajectory extrapolation [@fulvioTaskSpecificResponseStrategy2014], task switching [@sabahWhenLessMore2019], associative learning [@leeEvidentialDiversityIncreases2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice identity learning [@lavanEffectsHighVariability2019], simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018], sports training  [@greenPracticeVariabilityTransfer1995a; @northEffectConsistentVaried2019], and training on a complex video game [@seowTransferEffectsVaried2019]. See @czyzVariabilityPracticeInformation2021 or @ravivHowVariabilityShapes2022 for more detailed reviews.

Research on the effects of varied training typically manipulates variability in one of two ways. In the first approach, a high variability group is exposed to a greater number of unique instances during training, while a low variability group receives fewer unique instances with more repetitions. Alternatively, both groups may receive the same number of unique instances, but the high variability group's instances are more widely distributed or spread out in the relevant psychological space, while the low variability group's instances are clustered more tightly together. Researchers then compare the training groups in terms of their performance during the training phase, as well as their generalization performance during a testing phase. Researchers will usually compare the performance of the two groups during both the training phase and a subsequent testing phase. The primary theoretical interest is often to assess the influence of training variability on generalization to novel testing items or conditions. However, the test may also include some or all of the items that were used during the training stage, allowing for an assessment of whether the variability manipulation influenced the learning of the trained items themselves, or to make it easy to measure how much performance degrades as a function of how far away testing items are from the training items.

The influence of training variation has received a large amount of attention in the domain of sensorimotor skill learning. Much of this research has been influenced by the work of (@schmidtSchemaTheoryDiscrete1975), who proposed a schema-based account of motor learning as an attempt to address the longstanding problem of how novel movements are produced. Schema theory presumes that learners possess general motor programs for a class of movements (e.g. an underhand throw). When called up for use motor programs are parameterized by schema rules which determine how the motor program is parameterized or scaled to the particular demands of the current task. Schema theory predicts that variable training facilitates the formation of more robust schemas, which will result in improved generalization or transfer. Experiments that test this hypothesis are often designed to compare the transfer performance of a constant-trained group against that of a varied-trained group. Both groups train on the same task, but the varied group practices with multiple instances along some task-relevant dimension that remains invariant for the constant group. For example, studies using a projectile throwing task might assign participants to either constant training that practicing throwing from a single location, and a varied group that throws from multiple locations. Following training, both groups are then tested from novel throwing locations [@pigottMotorSchemaStructure1984; @willeyLimitedGeneralizationVaried2018; @pachecoLearningSpecificIndividual2018; @wulfEffectTypePractice1991].

One of the earliest, and still often cited investigations of Schmidt's benefits of variability hypothesis was the work of @kerrSpecificVariedPractice1978. Two groups of children, aged 8 and 12, were assigned to either constant or varied training of a bean bag throwing task. The constant group practiced throwing a bean-bag at a small target placed 3 feet in front of them, and the varied group practiced throwing from a distance of both 2 feet and 4 feet. Participants were blindfolded and unable to see the target while making each throw but would receive feedback by looking at where the beanbag had landed in between each training trial. 12 weeks later, all of the children were given a final test from a distance of 3 feet which was novel for the varied participants and repeated for the constant participants. Participants were also blindfolded for testing and did not receive trial by trial feedback in this stage. In both age groups, participants performed significantly better in the varied condition than the constant condition, though the effect was larger for the younger, 8-year-old children. This result offers a particularly compelling example of the merits of varied practice, given that the varied group was able to outperform the constant group even from the  home turf location where one may have expected the constatn group to have the strongest advantage. A similar pattern of results was observed in another study wherein varied participants trained with tennis, squash, badminton, and short-tennis rackets were compared against constant subjects trained with only a tennis racket [@greenPracticeVariabilityTransfer1995a]. One of the testing conditions had subjects repeat the use of the tennis racket, which had been used on all 128 training trials for the constant group, and only 32 training trials for the varied group. Nevertheless, the varied group outperformed the constant group when using the tennis racket at testing, and also performed better in conditions with several novel racket lengths. However, as is the case with many of the patterns commonly observed in the "benefits of variability" literature, the pattern wherein the varied group outperfroms the constant group even from the constants group's home turf has not been consistently replicated. One recent study attempted a near replication of the  Kerr & Booth study [@willeyLongtermMotorLearning2018], having subjects throw beanbags at a target, with the varied group training from positions (5 and 9 feet) on either side of the constant group (7 feet). This study did not find a varied advantage from the constant training position, though the varied group did perform better at distances novel to both groups. However, this study diverged from the original in that the participants were adults; and the amount of training was much greater (20 sessions with 60 practice trials each, spread out over 5-7 weeks).

Pitting varied against constant practice against each other on the home turf of the constant group provides a compelling argument for the benefits of varied training, as well as an interesting challenge for theoretical accounts that posit generalization to occur as some function of distance. However, despite its appeal this particular contrast is relatively uncommon in the literature. It is unclear whether this may be cause for concern over publication bias, or just researchers feeling the design is too risky. A far more common design is to have separate constant groups that each train exclusively from each of the conditions that the varied group encounters [@catalanoDistantTransferCoincident1984a; @chuaPracticeVariabilityPromotes2019;
@newellVariabilityPracticeTransfer1976; @moxleySchemaVariabilityPractice1979;
@mccrackenTestSchemaTheory1977], or for a single constant group to train from just one of the conditions experienced by the varied participants [@pigottMotorSchemaStructure1984; @rollerVariablePracticeLenses2001; @wrisbergTrainingProductionNovel1984; @wrisbergDevelopingCoincidentTiming1983]. A less common contrast places the constant group training in a region of the task space outside of the range of examples experienced by the varied group, but distinct from the transfer condition [@wrisbergVariabilityPracticeHypothesis1987; @wulfVariabilityPracticeImplicit1997]. Of particular relevance to the current work is the early study of @catalanoDistantTransferCoincident1984a, as theirs was one of the earliest studies to investigate the influence of varied vs. constant training on multiple testing locations of graded distance from the training condition. Participants were trained on coincident timing task, in which subjects observe a series of lightbulbs turning on sequentially at a consistent rate and attempt to time a button response with the onset of the final bulb. The constant groups trained with a single velocity of either 5,7,9, or 11 mph, while the varied group trained from all 4 of these velocities. Participants were then assigned to one of four possible generalization conditions, all of which fell outside of the range of the varied training conditions -- 1, 3, 13 or 15 mph. As is often the case, the varied group performed worse during the training phase. In the testing phase, the general pattern was for all participants to perform worse as the testing conditions became further away from the training conditions, but since the drop off in performance as a function of distance was far less steep for the varied group, the authors suggested that varied training induced a decremented generalization gradient, such that the varied participants were less affected by the change between training and testing conditions.

Benefits of varied training have also been observed in many studies outside of the sensorimotor domain.  @goodeSuperiorityVariableRepeated2008 trained participants to solve anagrams of 40 different words ranging in length from 5 to 11 letters, with an anagram of each word repeated 3 times throughout training, for a total of 120 training trials. Although subjects in all conditions were exposed to the same 40 unique words (i.e. the solution to an anagram), participants in the varied group saw 3 different arrangements for each solution-word, such as DOLOF, FOLOD, and OOFLD for the solution word FLOOD, whereas constant subjects would train on three repetitions of LDOOF (spread evenly across training). Two different constant groups were used. Both constant groups trained with three repetitions of the same word scramble, but for constant group A, the testing phase consisted of the identical letter arrangement to that seen during training (e.g. LDOOF), whereas for constant group B, the testing phase consisted of a arrangement they had not seen during training, thus presenting them with a testing situation similar situation to the varied group. At the testing stage, the varied group outperformed both constant groups, a particularly impressive result, given that constant group A had 3 prior exposures to the word arrangement (i.e. the particular permutation of letters) which the varied group had not explicitly seen. However varied subjects in this study did not exhibit the typical decrement in the training phase typical of other varied manipulations in the literature, and actually achieved higher levels of anagram solving accuracy by the end of training than either of the constant groups -- solving 2 more anagrams on average than the constant group. This might suggest that for tasks of this nature where the learner can simply get stuck with a particular word scramble, repeated exposure to the identical scramble might be less helpful towards finding the solution than being given a different arrangement of the same letters. This contention is supported by the fact that constant group A, who was tested on the identical arrangement as they experienced during training, performed no better at testing than did constant group B, who had trained on a different arrangement of the same word solution -- further suggesting that there may not have been a strong identity advantage in this task.

In the domain of category learning, the constant vs. varied comparison is much less suitable. Instead, researchers will 
typically employ designs where all training groups encounter numerous stimuli, but one group experiences a greater number of unique exemplars [@wahlheimMetacognitiveJudgmentsRepetition2012; @nosofskyModelguidedSearchOptimal2019; @doyleMetacognitiveMonitoringCategory2016; @hoschPriorExperienceVariability2023; @brunsteinPreparingNoveltyDiverse2011], or designs where the number of unique training exemplars is held constant, but one group trains with items that are more dispersed, or spread out across the category space [@posnerGenesisAbstractIdeas1968;  @homaCategoryBreadthAbstraction1976; @huHighvariabilityTrainingDoes2024; @bowmanTrainingSetCoherence2020; @maddoxStimulusRangeDiscontinuity2011].

Much of the earlier work in this sub-area trained subjects on artificial categories, such as dot patterns [@homaCategoryBreadthAbstraction1976; @posnerGenesisAbstractIdeas1968]. A seminal study by @posnerGenesisAbstractIdeas1968 trained participants to categorize artificial dot patterns, manipulating whether learners were trained with low variability examples clustered close to the category prototypes (i.e. low distortion training patterns), or higher-variability patterns spread further away from the prototype (i.e. high-distortion patterns). Participants that received training on more highly-distorted items showed superior generalization to novel high distortion patterns in the subsequent testing phase. It should be noted that unlike the sensorimotor studies discussed earlier, the @posnerGenesisAbstractIdeas1968 study did not present low-varied and high-varied participants with an equal number of training rathers, but instead had participants remain in the training stage of the experiment until they reached a criterion level of performance. This train-until-criterion procedure led to the high-variability condition participants tending to complete a larger number of training trials before switching to the testing stage. More recent work [@huHighvariabilityTrainingDoes2024], also used dot pattern categories, but matched the number of training trials across conditions. Under this procedure, higher-variability participants tended to reach lower levels of performance by the end of the training stage. The results in the testing phase were the opposite of @posnerGenesisAbstractIdeas1968,  with the low-variability training group showing superior generalization to novel high-distortion patterns (as well as generalization to novel patterns of low or medium distortion levels). However, whether this discrepancy is solely a result of the different training procedures is unclear, as the studies also differed in the nature of the prototype patterns used. @posnerGenesisAbstractIdeas1968  utilized simpler, recognizable prototypes (e.g., a triangle, the letter M, the letter F), while  @huHighvariabilityTrainingDoes2024 employed random prototype patterns.

Recent studies have also begun utilizing more complex or realistic sitmuli when assessing the influence of variability on category learning. @wahlheimMetacognitiveJudgmentsRepetition2012 conducted one such study. In a within-participants design, participants were trained on bird categories with either high repetitions of a few exemplars, or few repetitions of many exemplars. Across four different experiments, which were conducted to address an unrelated question on metacognitive judgements, the researchers consistently found that participants generalized better to novel species following training with more unique exemplars (i.e. higher variability), while high repetition training produced significantly better performance categorizing the specific species they had trained on. A variability advantage was also found in the relatively complex domain of rock categorization [@nosofskyModelguidedSearchOptimal2019]. For 10 different rock categories, participants were trained with either many repetitions of 3 unique examples of each category, or few repetitions of 9 unique examples, with an equal number of total training trials in each group (the design also included 2 other conditions less amenable to considering the impact of variation). The high-variability group, trained with 9 unique examples, showed significantly better generalization performance than the other conditions.

A distinct sub-literature within the category learning domain has examined how the variability or dispersion of the categories themselves influences generalization to ambiguous regions of the category space (e.g. the region between the two categories). The general approach is to train participants with examples from a high variability category and a low variability category. Participants are then tested with novel items located within ambiguous regions of the category space which allow the experimenters to assess whether the difference in category variability influenced how far participants generalize the category boundaries. @cohenCategoryVariabilityExemplar2001 trained subjects on two categories, one with much more variability than the other. In experiment 1, a low variability category composed of 1 instance was compared against a high-variability category of 2 instances in one condition, and 7 instances in another. In experiment 2 both categories were composed of 3 instances, but for the low-variability group the instances were clustered close to each other, whereas the high-variability groups instances were spread much further apart. Participants were tested on an ambiguous novel instance that was located in between the two trained categories. Both experiments provided evidence that participants were much more likely to categorize the novel middle stimulus into a category with greater variation.

Further observations consonant with the results of @cohenCategoryVariabilityExemplar2001  have since been observed in numerous investigations [@hahnEffectsCategoryDiversity2005; @perlmanFurtherAttemptsClarify2012;
@sakamotoPuttingPsychologyBack2008; @hsuEffectsGenerativeDiscriminative2010; but see @stewartEffectCategoryVariability2002]. The results of @sakamotoPuttingPsychologyBack2008 are noteworthy. They first reproduced the basic finding of participants being more likely to categorize an unknown middle stimulus into a training category with higher variability. In a second experiment, they held the variability between the two training categories constant and instead manipulated the training sequence, such that the examples of one category appeared in an ordered fashion, with very small changes from one example to the other (the stimuli were lines that varied only in length), whereas examples in the alternate category were shown in a random order and thus included larger jumps in the stimulus space from trial to trial. They found that the middle stimulus was more likely to be categorized into the category that had been learned with a random sequence, which was attributed to an increased perception of variability which resulted from the larger trial to trial discrepancies.

The work of @hahnEffectsCategoryDiversity2005, is also of particular interest to the present work. Their experimental design was similar to previous studies, but they included a larger set of testing items which were used to assess generalization both between the two training categories as well as novel items located in the outer edges of the training categories. During generalization testing, participants were given the option to respond with "neither", in addition to responses to the two training categories. The "neither" response was included to test how far away in the stimulus space participants would continue to categorize novel items as belonging to a trained category. Consistent with prior findings, high-variability training resulted in an increased probability of categorizing items in between the training categories as belong to the high variability category. Additionally, participants trained with higher variability also extended the category boundary further out into the periphery than participants trained with a lower variability category were willing to do. The author compared a variety of similarity-based models based around the Generalized Context Model [@nosofskyAttentionSimilarityIdentificationcategorization1986] to account for their results, manipulating whether a response-bias or similarity-scaling parameter was fit separately between variability conditions. No improvement in model fit was found by allowing the response-bias parameter to differ between groups, however the model performance did improvement significantly when the similarity scaling parameter was fit separately. The best fitting similarity-scaling parameters were such that the high-variability group was less sensitive to the distances between stimuli, resulting in greater similarity values between their training items and testing items. This model accounted for both the extended generalization gradients of the varied particpants, and also for their poorer performance in a recognition condition.

Variability has also been examined in the learning of higher-order linguistic categories [@perryLearnLocallyThink2010]. In nine training sessions spread out over nine weeks infants were trained on object labels in a naturalistic play setting. All infants were introduced to three novel objects of the same category, with participants in the "tight" condition being exposed to three similar exemplars of the category, and participants in the varied condition being exposed to three dissimilar objects of the same category. Importantly, the similarity of the objects was carefully controlled for by having a separate group of adult subjects provide pairwise similarity judgements of the category objects prior to the study onset. Multidimensional scaling was then performed to obtain the coordinates of the objects psychological space, and out of the 10 objects for each category, the 3 most similar objects were selected for the tight group and the three least similar objects for the varied group, with the leftover four objects being retained for testing. By the end of the nine weeks, all of the infants had learned the labels of the training objects. In the testing phase, the varied group demonstrated superior ability to correctly generalize the object labels to untrained exemplars of the same category. More interesting was the superior performance of the varied group on a higher order generalization task -- such that they were able to appropriately generalize the bias they had learned during training for attending to the shape of objects to novel solid objects, but not to non-solids. The tight training group, on the other hand, tended to overgeneralize the shape bias, leading the researchers to suggest that the varied training induced a more context-sensitive understanding of when to apply their knowledge.
